Management induced changes of soil organic carbon on global croplands

### 2022-07-20 SOCbudgetPaper/introduction.Rmd

Soil organic carbon (SOC), the amount of organic carbon stored in the Earth’s soil, constitutes the largest terrestrial organic carbon pool. It exceeds the carbon in the atmospheric and vegetation pools multiple times [@batjes_total_1996]. Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO~2~ in the atmosphere [@frielingstein_glob_carbon_2020; @minasny_soil_2017]. The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile [@batjes_total_1996; @sanderman_soil_2017].

@@

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs), Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions [@frielingstein_glob_carbon_2020]. While providing estimates of the magnitude of SOC losses due to land-cover change, most models lack a detailed representation of agricultural management. Earlier DGVM- and ESM-based assessments only considered changes in land cover, but ignored the removal of biomass at harvest [@strassmann_simulating_2008; @betts_climate_2015]. BKMs are designed to estimate LUC related emissions but often ignore changes in SOC due to climate change, CO~2~ fertilization and N deposition. Whereas BKM have largely improved in estimating additional emissions from wood harvest and shifting cultivation, state of the art models do not consider impacts of varying agricultural management [@frielingstein_glob_carbon_2020; @houghton_carbon_2012; @hansis_relevance_2015; @bastos_comparison_2021].

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance [e.g. @bondeau_modelling_2007; @lindeskog_implications_2013]. Pugh et al. [-@pugh_simulated_2015] explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with stylized scenarios rather than with historical management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historical LUC and the slow decline of SOC under cropland before a new equilibrium is reached.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time (e.g. growing seasons [@portmann_mirca2000global_2010], multiple cropping systems [@waha_multiple_2020], irrigation systems [@jagermeyr_water_2015]) or as stylized scenarios [e.g. @pugh_simulated_2015; @lutz_simulating_2019]. Herzfeld et al. [-@herzfeld_soc_2021] account for historical changes in fertilizer and manure inputs, residue removal rates and tillage systems and report SOC losses from cropland expansion over the period from 1700--2018 of 215 GtC. Within their stylized future management scenarios they find that none of the management aspects considered (residue management, no-tillage) can create a net carbon sink on current cropland areas under future climate change.

>>>>

Soil organic carbon (SOC), the amount of organic carbon stored in the Earth’s soil, exceeds the carbon in the atmospheric and vegetation pools multiple times [@batjes_total_1996]. Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO~2~ in the atmosphere [@frielingstein_glob_carbon_2020; @minasny_soil_2017]. The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile [@batjes_total_1996; @sanderman_soil_2017].

@@

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs), Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions [@frielingstein_glob_carbon_2020]. While providing estimates of the magnitude of SOC losses due to land-cover change, most models lack a detailed representation of agricultural management. Earlier DGVM- and ESM-based assessments only considered changes in land cover, but ignored the removal of biomass at harvest [@strassmann_simulating_2008; @betts_climate_2015]. BKMs are designed to estimate LUC related emissions and have largely improved in estimating additional emissions from wood harvest and shifting cultivation. However, state of the art models do not consider impacts of varying agricultural management [@frielingstein_glob_carbon_2020; @houghton_carbon_2012; @hansis_relevance_2015; @bastos_comparison_2021].

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance [e.g. @bondeau_modelling_2007; @lindeskog_implications_2013]. Pugh et al. [-@pugh_simulated_2015] explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with uniform scenario assumptions on management rather than with historical management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historical LUC and the slow decline of SOC under cropland before a new equilibrium is reached.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time [e.g. for growing seasons in @portmann_mirca2000global_2010; multiple cropping systems in @waha_multiple_2020; irrigation systems in @jagermeyr_water_2015] or as stylized (in the sense of uniform management assumptions) scenarios [e.g. @pugh_simulated_2015; @lutz_simulating_2019]. Herzfeld et al. [-@herzfeld_soc_2021] account for historical changes in fertilizer and manure inputs, residue removal rates and tillage systems and report SOC losses from cropland expansion over the period from 1700--2018 of 215 GtC. Within their stylized future management scenarios under future climate change they find that none of the management aspects considered (residue management, no-tillage) can create a positive SOC stock change on current cropland areas that counteracts the still continuing legacy flux from the initial land-use change.

### 2022-07-20 SOCbudgetPaper/methods.Rmd

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the Tier 2 modeling approach for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model [@parton_analysis_1987]. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and grasslands (referred to as natural vegetation in the following). Note that purpose grown annual forage crops are counted as cropland, whereas grazed pastures and mowed meadows (perennials) are assigned to natural vegetation. Carbon flows regarding livestock are split in the sense that only carbon originated from cropland feed sources are considered in this study.

@@

## Management scenario definitions {#sec:scenarios}

@@

>>>>

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the Tier 2 modeling approach for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model [@parton_analysis_1987]. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and grasslands (referred to as natural vegetation in the following). Forage crops are included within cropland, whereas pastures (including mowed meadows (perennials) and rangelands) are assigned to natural vegetation. Carbon flows connected to livestock are only considered in this study when they originate from cropland feed sources, while the manure originating from pasture biomass is disregarded, implicitly assuming that this manure is excreted or applied to pastures.

@@

## Scenario definitions {#sec:scenarios}

To highlight the impact of changing management effects and to assess the sensitivity of the model towards different initialization choices, we perform a set of scenario runs. In the following section we outline name and idea of these scenarios [for technical implementation see @karstens_mrsoil_2020].

@@

As outlined in Sect. \@ref(sec:initsoc) we assume to start in a steady-state SOC stocks for the start year of the spin-up in 1510 followed by a long spin-up period of (\textit{Initial-spinup1510}). As some SOC compartments decay over very long timescales, the initialization setting might strongly effect the overall outcome of SOC stocks and changes. Thus we conduct two counterfactual scenarios \textit{Initial-natveg} and \textit{Initial-eq}. Whereas in \textit{Initial-natveg} we assume SOC stocks to be in a steady-state SOC under potential natural vegetation for all land-use types in 1901, SOC stocks start in their land-use type specific steady state in 1901 for the \textit{Initial-eq} scenario. In that way, the two scenarios mark the two extreme cases: In \textit{Initial-eq} all legacy fluxes already appeared in 1901, whereas in \textit{Initial-natveg} all legacy fluxes before 1901 still have to appear. We additionally combined the counterfactual scenarios with the \textit{constManagement} scenario.

### 2022-07-20 SOCbudgetPaper/results.Rmd

Detailed results for the spatially explicit global SOC budget including intermediate results on input data as well as SOC stock results for all scenario runs can be found in the data repository from Karstens [-@karstens_model_2020]. In the following, the most important results [see @karstens_result_2020 for post-processing scripts] are summarized.

@@

```{r captionSOCdebt, echo=FALSE}
captionSOCdebt <- "Global total $\\Delta SOC$ and $\\Delta SOC$ change for the first 30 cm of the soil profile.
Panel (a) shows global $\\Delta SOC$ as the difference between $SOC$ under historical land use and $SOC$ under potential natural vegetation (PNV) in the year 2010 summed over all land-use types. Computing the difference between the $\\Delta SOC$ estimate for 2010 and for 1975 (b) depicts areas of soil depletion (SOC debt increase, red) and net-sequestration (SOC debt decline, blue)."
```

```{r SOCdebt, out.width = "9cm", fig.cap = captionSOCdebt, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/2panelfigure.png")
```
The spatial distribution of the total $\Delta SOC$ summed over all land-use types (Fig. \@ref(fig:SOCdebt)(a))  and its change from 1975 to 2010 (Fig. \@ref(fig:SOCdebt)(b)) reveals areas of SOC debt decline and increase. 
Regions with large cropland expansion (e.g. Brazil, Southeast Asia, Canada) continue to lose SOC, whereas regions with cropland reduction (and thus SOC restoration) or with accumulating cropland SOC can be found e.g. in highly productive areas of Europe and Central USA.

## Carbon flows in the agricultural system

```{r caption_FlowFig, echo=FALSE}
caption_FlowFig <- "Global carbon flows within the agricultural system for the year 2010 (in $\\unit{MtC}$). Carbon is first photosynthesised by crop plants and then used for livestock feed and various other usages subsumed under human demand. After accounting for losses within the agricultural system, there are three major C inputs to cropland SOC: manure, above- and below-ground residues. Large parts of C, however, are mineralized on the field before entering the soil. Additionally, C is transferred to and from the global agricultural soil stock via land-use change between cropland and natural vegetation. Finally, SOC is mineralized and flows back to the atmosphere."
```

@@

C is sequestered from the atmosphere via plant growth and allocated to different plant parts, which we aggregate to three pools (harvested organ, above- and below-ground residues). Whereas harvested organs as well as above ground-residues are taken (partially) from the field to be used for other purposes, below-ground residues ($729\,\unit{MtC}$ in 2010) are directly returned to the field. We divide crop biomass usage into feed usage and aggregate all other usage types (e.g. food, bioenergy and material) into a human demand category. Livestock feed demand for crop organ harvest and above-ground residues of $1136\,\unit{MtC}$ is roughly equal to the human demand of $1129\,\unit{MtC}$. Whereas large parts of feed intake are returned to the soils via manure (C input from manure at $384\,\unit{MtC}$), we assume the carbon demanded from humans (ending up as e.g. compost, night soil and sewage) is not returned to soils. Besides manure C and below-ground residues, above-ground residues form the largest C input to the soil with $1350\,\unit{MtC}$ returned to the fields in the year 2010. However, around 60% of this organic C decomposes before it is integrated into soils. Due to the different composition of organic C, proportionally more C enters the slow pool from manure than from crop residue. 
According to our model results, land-use change dynamics led to a C transfer from natural vegetation to cropland of $257\,\unit{MtC}$ in 2010. The agricultural system receives $4585\,\unit{MtC}$ assimilated by crop plants and releases $3554\,\unit{MtC}$ mostly through respiration. Accounting for SOC transfer and decomposition, the net SOC decrease of global cropland is around $33\,\unit{MtC}$ for the year 2010.

@@

>>>>

Detailed results for the spatially explicit global SOC budget including intermediate results on input data from manure and residues as well as SOC stock results for all scenario runs can be found in the data repository from Karstens [-@karstens_model_2020]. In the following, the most important results [see @karstens_result_2020 for post-processing scripts] are summarized.

@@

```{r captionSOCdebt, echo=FALSE}
captionSOCdebt <- "Global total $\\Delta SOC$ and $\\Delta SOC$ change for the first 30 cm of the soil profile. Panel (a) shows global $\\Delta SOC$ as the difference between $SOC$ under historical land use and $SOC$ under potential natural vegetation (PNV) in the year 2010 summed over all land-use types. Computing the difference between the $\\Delta SOC$ estimate for 2010 and for 1975 (b1) depicts areas of SOC depletion (SOC debt increase, red) and SOC accumulation (SOC debt decline, blue). Panel (b2) shows the of land-use change (LUC) induced change of ∆SOC between 1975 and 2010, whereas panel (b3) depicts the change due to changing agricultural management (MAN)."
```

```{r SOCdebt, out.width = "9cm", fig.cap = captionSOCdebt, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/SpatialDebtFigure.png")
```
The spatial distribution of the total $\Delta SOC$ summed over all land-use types (Fig. \@ref(fig:SOCdebt)(a))  and its change from 1975 to 2010 (Fig. \@ref(fig:SOCdebt)(b1)) reveals areas of SOC debt decline and increase. 
Regions with large cropland expansion (e.g. Brazil, Southeast Asia, Canada) continue to lose SOC, whereas regions with cropland reduction (and thus SOC restoration) or with accumulating cropland SOC can be found e.g. in highly productive areas of Europe and Central USA.

## Carbon flows in the cropland system

```{r caption_FlowFig, echo=FALSE}
caption_FlowFig <- "Global carbon flows within the cropland system for the year 2010 (in $\\unit{MtC}$). Carbon is first photosynthesised by crop plants and then used for livestock feed and various other usages subsumed under human demand. After accounting for losses within the cropland system, there are three major C inputs to cropland SOC: manure, above- and below-ground residues. Large parts of C, however, are mineralized on the field before entering the soil. Additionally, C is transferred to and from global cropland soil stock via land-use change between cropland and natural vegetation. Finally, SOC is mineralized and flows back to the atmosphere."
```

@@

C is removed from the atmosphere via plant growth and allocated to different plant parts, which we aggregate to three pools (harvested organ, above- and below-ground residues). Whereas harvested organs as well as above ground-residues are taken (partially) from the field to be used for other purposes, below-ground residues ($729\,\unit{MtC}$ in 2010) are directly returned to the field. We divide crop biomass usage into feed usage and aggregate all other usage types (e.g. food, bioenergy and material) into a human demand category. Livestock feed demand for crop organ harvest and above-ground residues of $1136\,\unit{MtC}$ is roughly equal to the human demand of $1129\,\unit{MtC}$. Whereas large parts of feed intake are returned to the soils via manure (C input from manure at $384\,\unit{MtC}$), we assume the carbon demanded from humans (ending up as e.g. compost, night soil and sewage) is not returned to soils. Besides manure C and below-ground residues, above-ground residues form the largest C input to the soil with $1350\,\unit{MtC}$ returned to the fields in the year 2010. However, around 60% of this organic C decomposes before it is integrated into soils. Due to the different composition of organic C, proportionally more C enters the slow pool from manure than from crop residue. 
According to our model results, land-use change dynamics led to a C transfer from natural vegetation to cropland of $257\,\unit{MtC}$ in 2010. The cropland system receives $4585\,\unit{MtC}$ assimilated by crop plants and releases $3554\,\unit{MtC}$ mostly through respiration. Accounting for SOC transfer and decomposition, the net SOC decrease of global cropland is around $33\,\unit{MtC}$ for the year 2010.

@@

Using the constManagement results, that only include land-use change (LUC) related changes of the SOC debt between 1975 and 2010, we are able to subtract the LUC effect from the overall SOC debt change within the histManagement results. The remaining effect can be attributed to the changing agricultural management (MAN) as other drivers such as climatic effects have been already canceled out by taking the difference to a PNV reference state while calculating $\Delta SOC$. 
The increasing SOC debt on global cropland are primarily caused by LUC (red areas in Fig. \@ref(fig:SOCdebt)(b2)). Deteriorated management also contributed to increasing SOC debt in parts of Sub-saharan Africa and Central Asia. In contrast, agricultural management has led to an decrease in SOC debt in the USA, Europe, as well as in parts of China and India (blue areas in Fig. \@ref(fig:SOCdebt)(b3)), which is not visible in the total $\Delta SOC$ change as LUC was happening at the same time.

Our sensitivity analysis shows that the management impact is robust to the initialization of SOC stocks (see Fig. \@ref(fig:SOCinit) and Sect. \@ref(sec:scenarios)) with around $2.15\,\unit{GtC}$ difference in SOC debt between the histManagement and the constManagement scenario. However, the SOC debt and SOC debt change varies with the different initialization choices. Whereas the default assumption (Initial-spinup1510) shows a $\Delta SOC$ of $39.6\,\unit{GtC}$ for the year 2010, the Initial-natveg scenario with high legacy fluxes to come only has a $\Delta SOC$ of $33.3\,\unit{GtC}$ and the Initial-eq scenarios with only little legacy fluxes left already a $\Delta SOC$ of $50.7\,\unit{GtC}$.
